Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-05-16 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Edits to ms

We performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds with eaach species or mixture replicated 5 times at each moisture level. Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling. This design allows fire behaviour to reach constant flame spread rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers   equally spaced along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction. If a species at a specific moisture level failed to ignite after three attempts, no further attempts were made.  A timer was used to record time to ignition and duration of flaming combustion. The flammability trials followed the methodology detailed in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.

@@

The effect of time since wetting on flammability parameters was modelled using linear regression with the interaction between time since wetting and genus as fixed effects and average vapor pressure deficit in the burning room during that trial as a nuisance covariate. Differences across species were tested with post-hoc Tukey tests. A binomial logistic regression was fitted to the probability of ignition to obtain moisture of extinction.

@@

There are marked differences in the extent of moisture species retain initially following wetting and draining by gravity (all p < 0.0001, Fig. \ref{fig1-drydown} and Table \ref{tab1-drydown}). \emph{Q. kelloggii}, the only broadleaf, is the species that retains the most moisture but has had a rapid dessication rate (. Calculating estimated marginal means following the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}.  There are also differences in desiccation rates across the eight species and estimated mixed linear model slopes discriminated species into three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}.

\begin{figure}[H]
  \centering
  \label{fig1-drydown}
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption[Dry down curves for eight litter types.]{Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis. Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing. Colors indicate genera and there were significant differences in both }
\end{figure}

\begin{table}[H]
  \caption{Mixed model results for moisture content as a function of time since wetting. Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
  \label{tab1-drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/drydown-tab.ltx}
\end{table}

Across these eight species specific leaf area had a positive effect on maximum water retention (Fig. \ref{maxmc}, linear model p = 0.003 and litter bulk density had a negative effect on desiccation rate (Fig. \ref{fig-bd-di}, linear model p = 0.029). Anova tables are shown in in Fig. \ref{tab-mc-anova} and Fig. \ref{tab-di-anova}.

\begin{figure}[H]
  \centering
  \label{fig-maxmc-di}
\includegraphics[width=16cm]{figs_tables/fig3-SLA-maxMC.pdf}
\caption[Maximum water retention by specific leaf area]{Maximum water retention by specific leaf area (SLA) across eight species.  TODO }
\end{figure}

\begin{table}[H]
  \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density.}
  \label{tab2-mc-anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/mc-anova.ltx}
\end{table}

\begin{figure}[H]
  \centering
  \label{fig-bd-di}
\includegraphics[width=16cm]{figs_tables/fig4_di_by_bd.pdf}
\caption[Litter density effect on dessication rate.]{  TODO }
\end{figure}

\begin{table}[H]
  \caption{Linear model results for estimated dessication rate ($hr^{-1}$) as a function of specific leaf area and litter bulk density.}
  \label{tab2-di-anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/di-anova.ltx}
\end{table}

@@

Time to ignition and spread rate were modeled against time since wetting and using genus as a fixed effect.

[TODO]?  redo in terms of groups.

of time since wetting and genus on time to ignition and spread rate (F10, 171=24.51, p < 2.2e-16 and F10, 171=55. 85, p<2.2e-16 respectively; Figures 2.3 and 2.4). There were also significant interactions of time since wetting and genus on the response variables (time to ignition and spread rate; both p < 2.2e-16). For time since ignition, the negative relationship is driven by Sequoiadendron and Calocedrus, as evidenced by the steeper slopes of the curves. However, if I removed Sequoiadendron and Calocedrus from the models, the interaction of time since wetting and genus disappears for time to ignition (p=0.1643) but is still significant for spread rate (p = 2.05e-6). After drying for 72 hours, Q. kelloggi is always more flammable and more responsive to changes in moisture content, while C. decurrens and S. giganteum are not (Figures 2.3 and 2.4).

@@

>>>>

We performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds with each species or mixture replicated 5 times at each moisture level. Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling. This design allows fire behaviour to reach constant flame spread rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers   equally spaced along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction. If a species at a specific moisture level failed to ignite after three attempts, no further attempts were made.  A timer was used to record time to ignition and duration of flaming combustion. The flammability trials followed the methodology detailed in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.

@@

Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing and therefore species within a genus tended to behave similarly with
the two scale-leaved genera in the Cupressaceae, \emph{Calocedrus} and
\emph{Sequoiadendron} also behaving similarly. Therefore, we used taxon
(\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as a fixed in
our models of moisture effects on flammability parameters. The effect of time
since wetting on flammability parameters was modelled using linear regression
with the interaction between time since wetting and genus as fixed effects and
average vapor pressure deficit in the burning room during that trial as a
nuisance covariate. Differences across species were tested with post-hoc Tukey
tests. A binomial logistic regression was fitted to the probability of ignition
to obtain moisture of extinction.

@@

There are marked differences in the extent of moisture species retain initially following wetting and draining by gravity (all p < 0.0001, Fig. \ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only broadleaf, is the species that retains the most moisture initially.  \emph{Quercus}, however, like the pines has a rapid dessication rate (Fig. \ref{fig:drydown-logged}). Calculating estimated marginal means following the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa} (SI 1).  There are also differences in desiccation rates across the eight species and estimated mixed linear model slopes discriminated species into three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI 2).

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption[Dry down curves for eight litter types.]{Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis. Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing. Colors indicate genera and there were significant differences in both }
 \label{fig:drydown}
\end{figure}

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig2_drydown-curves_logged.pdf}
\caption[Semi-log scale dry down curves for eight litter types.]{Lines show best-fit lines fit to the natural log of moisture content on a dry mass basis.  The slope of the line shows the dessication rate ($hr^{-1}$). Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing. Colors indicate genera and there were significant differences in both }
 \label{fig:drydown-logged}
\end{figure}

\begin{table}[H]
  \caption{Mixed model results for moisture content as a function of time since wetting. Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
  \label{tab:drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/drydown-tab.ltx}
\end{table}

Across these eight species specific leaf area had a positive effect on maximum water retention (Fig. \ref{fig:maxmc-di}, linear model p = 0.003) and litter bulk density had a negative effect on desiccation rate (Fig. \ref{fig:bd-di}, linear model p = 0.029). Anova tables are shown in in Fig. \ref{tab-mc-anova} and Fig. \ref{tab:di-anova}.

\begin{figure}[H]
  \centering
\includegraphics[width=16cm]{figs_tables/fig3-SLA-maxMC.pdf}
\caption[Maximum water retention by specific leaf area]{Maximum water retention by specific leaf area (SLA) across eight species.  TODO }
  \label{fig:maxmc-di}
\end{figure}

\begin{table}[H]
  \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density.}
  \label{tab2:mc-anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/mc-anova.ltx}
\end{table}

\begin{figure}[H]
  \centering
\includegraphics[width=16cm]{figs_tables/fig4_di_by_bd.pdf}
\caption[Litter density effect on dessication rate.]{  TODO }
  \label{fig:bd-di}
\end{figure}

\begin{table}[H]
  \caption{Linear model results for estimated dessication rate ($hr^{-1}$) as a function of specific leaf area and litter bulk density.}
  \label{tab2:di-anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/di-anova.ltx}
\end{table}

@@

Time to ignition and spread rate were modeled against time since wetting and using genus as a fixed effect and species as a random effect.  Time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by genus (Fig. \ref{fig:spreadrate}, Table \ref{tab:spreadrate-anova}). The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting. At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had much drying litter than did the \emph{Quercus}, yet the spread rates were slow.  \emph{Quercus} had the steepest response to time since wetting (Fig. \ref{fig5:spreadrate}).

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig5_spread_by_time.pdf}
\caption[Flame spread rate by time since wetting]{TODO }
  \label{fig:spreadrate}
\end{figure}

\begin{table}[H]
  \caption[Mixed linear model of flame spread rate.]{Mixed model results for flame spread rate  as a function of time since wetting with genus as a fixed effect and species as a nested random effect. Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
  \label{tab:spreadrate-anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/spreadrate-anova.ltx}
\end{table}

@@

NOTE: P lambertiana higher water absorption, presumably due to higher SLA.
